---
title: "Tidy Tuesday 2019 week 44: NYC squirrel census"
date: "2019-10-29"
output:
    rmarkdown::html_document:
      code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = F, warning = F, message = F)
library(tidyverse)
library(leaflet)
htmltools::tagList(rmarkdown::html_dependency_font_awesome())
```


```{r dataload}
nyc_squirrels <- readr::read_csv("https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2019/2019-10-29/nyc_squirrels.csv")
```

This [Tidy Tuesday](https://github.com/rfordatascience/tidytuesday/blob/master/data/2019/2019-10-29) explores the [squirrel census!](https://www.thesquirrelcensus.com) I took the opportunity to try to learn a little [Leaflet.](https://leafletjs.com)

```{r}
# date string to date time
nyc_squirrels$date_2 <- lubridate::mdy(nyc_squirrels$date)

# above ground FALSE to 0 and character to numeric
nyc_squirrels$above_ground_sighter_measurement <- ifelse(nyc_squirrels$above_ground_sighter_measurement == "FALSE", 0, nyc_squirrels$above_ground_sighter_measurement) %>% 
  as.numeric() 

# max(nyc_squirrels$above_ground_sighter_measurement, na.rm= T)
# summary(nyc_squirrels$above_ground_sighter_measurement)
nyc_squirrels <- nyc_squirrels %>% 
  filter(!is.na(above_ground_sighter_measurement))

# hex map
hex_map<- ggplot(nyc_squirrels, 
                 aes(x = lat, y = long, 
                     color = above_ground_sighter_measurement)) +
  geom_hex()

# set leaflet map

# all heights
pal <- colorBin(palette = "RdYlBu", 
                    domain = nyc_squirrels$above_ground_sighter_measurement, 
                    bins = 10,
                    reverse = TRUE)

nyc_squirrels_map <- leaflet(nyc_squirrels) %>% 
  addProviderTiles(providers$CartoDB) %>% 
  addCircleMarkers(radius = 2, 
                   color = ~pal(above_ground_sighter_measurement)) %>%
  addLegend("bottomright", 
            pal = pal, 
            values = ~above_ground_sighter_measurement, 
            title = "Squirrel height above ground") %>% 
  addScaleBar("bottomleft")

# less than 50 and not ground
nyc_squirrels_l50 <- nyc_squirrels %>% 
  filter(above_ground_sighter_measurement > 0) %>%
  filter(above_ground_sighter_measurement < 50)

pal <- colorBin(palette = "RdYlBu", 
                    domain = nyc_squirrels_l50$above_ground_sighter_measurement, 
                    bins = 10,
                    reverse = TRUE)

nyc_squirrels_l50_map <- leaflet(nyc_squirrels_l50) %>% 
  addProviderTiles(providers$CartoDB) %>% 
  addCircleMarkers(radius = 2, 
                   color = ~pal(above_ground_sighter_measurement)) %>%
  addLegend("bottomright", 
            pal = pal, 
            values = ~above_ground_sighter_measurement, 
            title = "Squirrel height above ground") %>% 
  addScaleBar("bottomleft")


# above 50
nyc_squirrels_g50 <- nyc_squirrels %>% 
  filter(above_ground_sighter_measurement > 50)

pal <- colorBin(palette = "RdYlBu", 
                    domain = nyc_squirrels_g50$above_ground_sighter_measurement, 
                    bins = 10,
                    reverse = TRUE)

nyc_squirrels_g50_map <- leaflet(nyc_squirrels_g50) %>% 
  addProviderTiles(providers$CartoDB) %>% 
  addCircleMarkers(radius = 2, 
                   color = ~pal(above_ground_sighter_measurement)) %>%
  addLegend("bottomright", 
            pal = pal, 
            values = ~above_ground_sighter_measurement, 
            title = "Squirrel height above ground") %>% 
  addScaleBar("bottomleft")
```

The first map shows all squirrels with given distance above ground at which they were sighted. NA's were filtered out. According to [the data dictionary](https://github.com/rfordatascience/tidytuesday/tree/master/data/2019/2019-10-29#data-dictionary), fields were populated with a value of “FALSE” if the squirrel was on the ground plane so these values were changed to zeros. This map is mostly filled with dark blue, the color of ground-dwelling squirrels.

`r nyc_squirrels_map`

Next, ground plane was filtered out and only values below 50 are shown. This map is a little more manageable.

`r nyc_squirrels_l50_map`


The final map shows values above 50. There are only a handful of such points.

`r nyc_squirrels_g50_map`

I couldn't find any information about the units for the measurement, so I'm not sure if the upper range makes any sense...

<hr />
